\pkg{RProtoBuf}: Efficient Cross-Language Data Serialization in R

### 2014-01-15 papers/jss/article.Rnw
some tweaks to intro
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/rprotobuf@789 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\section{Introduction: Friends don't let friends use CSV}

@@

Traditionally, scientists and statisticians often use character seperated
text formats such as \texttt{CSV} \citep{shafranovich2005common} to 
export and import data. However, anyone who has ever used \texttt{CSV} will have
noticed that this method has many limitations: it is restricted to tabular 
data, lacks type-safety, and has limited precision for numeric values.
Moreover, ambiguities in the format itself frequently cause problems. 
For example, conventions on which characters used as seperator and decimal
point vary by country.
\emph{Extensible Markup Language} (\texttt{XML}) is another
well-established and widely-supported format with the ability to define
just about any arbitrarily complex schema \citep{nolan2013xml}. However,
it pays for this complexity with comparatively large and verbose messages,
and added complexitiy at the parsing side (which are somewhat mitigated
by the availability of mature libraries and parsers). Because \texttt{XML}
is text based and has no native notion of numeric types or arrays, it 
usually not a very practical format to store numeric datasets as they appear
in statistical applications.
A more modern, widely used format is \emph{JavaScript Object Notation}
(\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and used mostly on the web. \texttt{JSON} natively
supports arrays and distinguishes 4 primitive types: numbers, strings, 
booleans and null. However, because it is a text-based format, numbers are
stored as human-readable decimal notation which is somewhat inefficient and
leads to loss of type (double vs integer) and precision. Several R packages
implement functions to parse and generate \texttt{JSON} data from R objects. 
A number of \texttt{JSON} dialects has been proposed, such as \texttt{BSON} and 
\texttt{MessagePack} which both add binary support. However, these derivatives
are not compatible with existing JSON software, and have not been widely adopted.

\subsection{Why Protocol Buffers}

In 2008, Google released an open source version of Protocol Buffers: the data 
interchange format that was designed and used for their internal infrastructure.
Google officially provides high quality parsing libraries for \texttt{Java}, 
\texttt{C++} and \texttt{Python}, and community developed open source implementations
are available for many other languages. 
Protocol Buffers take a quite different approach from many other popular formats.
They offer a unique combination of features, performance, and maturity that seems
particulary well suited for data-driven applications and numerical computing.
Protocol Buffers are a binary format that natively supports all common primitive types
found in modern programming languages. The advantage of this is that numeric values
are serialized exactly the same way as they are stored in memory. Therefore there is
no loss of precision, no overhead, and parsing messages is very efficient: the system can 
simply copy bytes to memory without any further processing. 
But the most powerful feature of protocol buffers is that it decouples the content
from the structure using a schema, very similar to a database. This further increases
performance by eliminating redundancy, while at the same time providing foundations
for defining an \emph{Interface Description Language}, or \emph{IDL}.
Many sources compare data serialization formats and show Protocol Buffers compare 
very favorably to the alternatives; see \citet{Sumaray:2012:CDS:2184751.2184810} 
for one such comparison.

@@

This paper describes an R interface to protocol buffers.
The rest of the paper is organized as follows. Section~\ref{sec:protobuf}
provides a general overview of Protocol Buffers.
Section~\ref{sec:rprotobuf-basic} describes the interactive R interface
provided by \CRANpkg{RProtoBuf} and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
describes the implementation details of the main S4 classes making up this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between R and other languages.  Section~\ref{sec:evaluation} introduces a
general R language schema for serializing arbitrary R objects and evaluates
it against R's built-in serialization.  Sections~\ref{sec:opencpu}
and \ref{sec:mapreduce} provide real-world use cases of \CRANpkg{RProtoBuf}
in web service and MapReduce environments, respectively, before
Section~\ref{sec:summary} concludes.

>>>>

\section{Introduction} % TODO(DE) More sober: Friends don't let friends use CSV}

@@

%\paragraph*{Friends don't let friends use CSV!}
Data analysts and researchers often use character seperated text formats such
as \texttt{CSV} \citep{shafranovich2005common} to export and import
data. However, anyone who has ever used \texttt{CSV} files will have noticed
that this method has many limitations: it is restricted to tabular data,
lacks type-safety, and has limited precision for numeric values.  Moreover,
ambiguities in the format itself frequently cause problems.  For example,
conventions on which characters is used as seperator or decimal point vary by
country.  \emph{Extensible Markup Language} (\texttt{XML}) is another
well-established and widely-supported format with the ability to define just
about any arbitrarily complex schema \citep{nolan2013xml}. However, it pays
for this complexity with comparatively large and verbose messages, and added
complexitiy at the parsing side (which are somewhat mitigated by the
availability of mature libraries and parsers). Because \texttt{XML} is text
based and has no native notion of numeric types or arrays, it usually not a
very practical format to store numeric datasets as they appear in statistical
applications.  A more modern, widely used format is \emph{JavaScript Object
  Notation} (\texttt{JSON}), which is derived from the object literals of
\proglang{JavaScript}, and used increasingly on the world wide web. \texttt{JSON} natively
supports arrays and distinguishes 4 primitive types: numbers, strings,
booleans and null. However, as it too is a text-based format, numbers are
stored as human-readable decimal notation which is somewhat inefficient and
leads to loss of type (double versus integer) and precision. Several R packages
implement functions to parse and generate \texttt{JSON} data from R objects.
A number of \texttt{JSON} variants has been proposed, such as \texttt{BSON}
and \texttt{MessagePack} which both add binary support. However, these
derivatives are not compatible with existing JSON software, and have not seen
wide adoption.

%\paragraph*{Enter Protocol Buffers:}
In 2008, and following several years of internal use, Google released an open
source version of Protocol Buffers. It provides data  
interchange format that was designed and used for their internal infrastructure.
Google officially provides high-quality parsing libraries for \texttt{Java}, 
\texttt{C++} and \texttt{Python}, and community-developed open source implementations
are available for many other languages. 
Protocol Buffers take a quite different approach from many other popular formats.
They offer a unique combination of features, performance, and maturity that seems
particulary well suited for data-driven applications and numerical computing.
Protocol Buffers are a binary format that natively supports all common primitive types
found in modern programming languages. A key advantage is that numeric values
are serialized exactly the same way as they are stored in memory. There is
no loss of precision, no overhead, and parsing messages is very efficient: the system can 
simply copy bytes to memory without any further processing. 
But the most powerful feature of Protocol Buffers is that it decouples the content
from the structure using a schema, very similar to a database. This further increases
performance by eliminating redundancy, while at the same time providing foundations
for defining an \emph{Interface Description Language}, or \emph{IDL}.
Many sources compare data serialization formats and show Protocol Buffers compare 
very favorably to the alternatives; see \citet{Sumaray:2012:CDS:2184751.2184810} 
for one such comparison.
% TODO(DE): Mention "future proof" forward compatibility of schemata

@@

This paper describes an R interface to Protocol Buffer, 
and is organized as follows. Section~\ref{sec:protobuf}
provides a general overview of Protocol Buffers.
Section~\ref{sec:rprotobuf-basic} describes the interactive R interface
provided by \CRANpkg{RProtoBuf} and introduces the two main abstractions:
\emph{Messages} and \emph{Descriptors}.  Section~\ref{sec:rprotobuf-classes}
describes the implementation details of the main S4 classes making up this
package.  Section~\ref{sec:types} describes the challenges of type coercion
between R and other languages.  Section~\ref{sec:evaluation} introduces a
general R language schema for serializing arbitrary R objects and evaluates
it against R's built-in serialization.  Sections~\ref{sec:opencpu}
and \ref{sec:mapreduce} provide real-world use cases of \CRANpkg{RProtoBuf}
in web service and MapReduce environments, respectively, before
Section~\ref{sec:summary} concludes.
